Predicting the conformational preferences of proteins using a physics-based free energy method

### 2013-01-13 Paper/paper.tex
Cleaned up intro and first section of results.

\author{Arijit~Roy}
\author{Justin~L.~MacCallum}
\author{Alberto~Perez}
\author{Ken~A.~Dill}
\affil{Laufer Center for Physical and Quantitative Biology\\
    Stony Brook University\\
    Stony Brook, NY 11794-5252.}

\title{Physics-based protein structure prediction and design using the confinement method}

@@

In computational structural biology there are numerous cases where the free energy difference
between two well defined 
states plays an important role. Examples range from small to large conformational changes of proteins due to
ligand binding, change of pH or other conditions \cite{Meirovitch2007}, \cite{Chipot2007}, \cite{Jorgensen2004}. 
Free energy also plays an important role in the case of protein
folding. Ground breaking work of Christian B. Anfinsen and coworkers showed that the native
structures of small globular proteins have a unique, thermodynamically stable native structure with
their conformation at the global free energy minimum \cite{Anfinsen1973}. Regardless of the starting
point or even after unfolding 
it by changing different condition most proteins will go back to the native state after
reinstituting native like conditions. Often, the changes in free energy between different
conformations is small, but nonetheless crucial in favoring the native state
 since defects in protein folding may be the molecular cause of a range of human genetic disorders.
 For example a misfolded protein known as a
prion can misfold correctly folded proteins when entering a healthy organism. Thus it is important for 
protein to have its native structure. As protein native structure stays in the global free energy minimum, 
\Arijit{free energy can be used as a scale for accurate identification of misfolded state from the protein native state.}
\Alberto{I'm not sure I understand this here} 
These facts further give free energy a special importance in structural biology.

However, the theoretical calculation of conformational free energy change becomes difficult if the
states of interest are very different \cite{Meirovitch2007}. In such scenario timescales involved in
such conformational transitions may
be beyond the sampling ability of classical molecular dynamics simulation.  Generally, in order to
calculate the free energy a pathway or reaction coordinate is needed with several intermediate structures.
Methods like umbrella sampling \cite{Torrie1977}, thermodynamic integration \cite{Tironi1994} along with classical MD can suffer from
overlap problem and requires a larger computational effort. The idea of a reaction coordinate became
even more
complicated in the case of protein folding.  As free energy landscape for protein folding 
is rough \cite{Dill1997}, \cite{Dill2008}, the molecule 
can get trapped in an energy well during conformational sampling. Even with 
advanced methods like replica exchange molecular dynamics, the three dimensional structure may be
trapped in a local minima for a considerable time, giving rise to nonconverged distributions where
misfolded states might be selected as native like
% (give the example of human pin1ww domain from Benout Roux and Klaus Schultan).

The calculation of free energy have been successfully attempted by a number of groups \cite{Meirovitch2007}, 
\cite{Ytreberg2006} - \cite{Zheng2008}. In recent years methods like the Reference system method \cite{Ytreberg2006}, 
Decativated Morphing \cite{Park2008}, Orthogonal Space Random Work \cite{Zheng2008} or the Confinement 
Method \cite{Tyka2006}, \cite{Cecchini2009} to name a few have been used to calculate free energy
differences.
In this work we have applied the confinement method which was originally devoled by Tyka et. al. \cite{Tyka2006} and
Cecchini et. al. \cite{Cecchini2009}. This method relies on the fact that the free energy difference is a state function and 
thus it is independent of the sampling path. This approach uses a thermodynamic cycle where first, non-harmonic degrees of
freedom are removed by applying a series of restraints to the system. Then the free energy of the
remaining harmonic system is obtained using a normal mode calculation and combined with the
non-harmonic part to give the total absolute free energy.  Previously, this method was applied to
calculate the conformational free energy of a 16 amino acid residue peptide, known as BHP.  We have
first reproduced that known result. Next we have applied this method to larger proteins for the
first time. The free energy difference of pairs of conformations in proteins with similar sequence but completely
different fold was calculated. \Alberto{i made this part too long, have to summarize here and define
better inside text?-->} Encouraged by the result, we applied the method to help tackle the
problem of protein conformation ranking. We did this by selecting different targets from the
Critical Assessment of Structure Prediction (CASP) event. In this event, different research groups
try to find the 3D structure of a protein given a sequence in a blind way using their methodologies. Upon comparing
with the real structure it is often seen that several groups are able to generate better structures
than the ones they submit, but are not able to identify them. This is known as the ranking problem,
and this method is able to correctly rank structures in the majority of cases we have tried.
 Finally, we have been exploring the ability of working on a per residue basis to identify which
 residues are the major players in the stabilization/destabilization of a pair of conformations. Our
 initial experience makes us thinkg that this method might be helpful in identifying key residues
 for protein design.

%of protein can be obtained to atomic resolution by X-ray crystallography or NMR. Unfortunately, the
%sheer amount of time and economic investment for structure determination cannot keep up with the
%rate at which proteins of unknown structure are sequenced. Therefore, computational models can be a
%good way to bridge the increasing gap between sequence and structure. In this respect CASP
%(Critical Assessment of Structure Prediction) has played a crucial role which is a community wide
%experiments for the protein structure prediction.

@@

\subsection{The confinement method produces correct results in control experiments}

\Alberto{ This is benchmarking, good internal, but not relevant. Should go to SI,out, or summarized
    in one sentence:  "After reproducing the results of Cecchini and coworkers we were interested in
the ability to apply this method to larger systems"}

\Alberto{As a first step, we performed several control experiments to verify that our implementation of the
confinement method produces results compatible with previous calculations reported in the
literature.

The method has previously been applied to a 16 amino acid residue $\beta$-hairpin from
protein G, known as BHP \cite{Cecchini2009}. We calculated the free energy difference between two different
conformations of the peptide: (1) the native conformation, called bhp1, with a two-stranded
$\beta$-sheet; and (2) a conformation, called bhp3, which has a three-stranded $\beta$-sheet.
Analysis of long (4 $\mu$s) equilibrium simulations \cite{Cecchini2009},\cite{Krivov2004}  shows that bhp1 is the more favorable
configuration by 1.8 kcal/mol. Using the confinement method, we obtain a value 1.7 kcal/mol, which
is in good agreement with the equilibrium simulations and with previous calculations using the
confinement method \cite{Cecchini2009} (see Supporting Information for further details).}

Previous applications of the confiment method have focused on relatively short peptides, up to $17$
residues in length. In this work we apply the confinement method to
larger proteins. On that direction, we first test the confinement method for a chamelon sequence and 
found that it can correctly pick out the preferential structure of the chamelon sequence. Other 
applications we explore in the present work is the re-scoring, or
metaprediction, of structures submitted during the Critical Assessment of Structure Prediction
(CASP) experiment (described in detail later). We computed the relative free energies of predictions
submitted during CASP9, with the expectation that the most native-like prediction will have
the lowest free energy. As a control, for several targets we also calculated the relative free
energy of the experimentally determined structure. \Alberto{should not start with results yet, first
we are describing what we are going to show them} As
Table~\ref{table:casp_control} shows, in most cases, the confinement method correctly assigns a
lower free energy to the experimentally determined structure than to any of the decoys. 
Most interestingly, we decomposed the free energy into its per residue component, which help 
us to identify the residues which stabilize or destabilize a particular conformation of the protein.
%Although differentiating between experimentally determined structures and computer generated predictions is
%not a stringent test of a scoring method---for example, examining the residue-residue packing can
%easily distinguish between the two [ref Rosetta Holes]---the results none the less serve as a useful
%"sanity check".

\begin{table}
\begin{center}
\caption{The confinement method assigns a more favorable free energy to the experimentally
determined structure than to computer-generated predictions. For each target, we examined as many as
five predictions submitted by CASP participants. Positive $\Delta\Delta G$ values indicate that the
experimental structure is predicted to be more favorable than any of the decoys.}
\label{table:casp_control}
\begin{tabular}{l l l}\hline
    CASP Target  & PDB Identifier & $\Delta\Delta G_{native \to best decoy}$ (kcal/mol) \\ \hline
     T0531       &    2KJX        &          $11.15 \pm 0.70$ \\ \hline
     T0538       &    2L09        &          $-3.00 \pm 0.47$ \\ \hline
     T0540       &    3MX7        &          $16.94 \pm 0.49$ \\ \hline
     T0559       &    2L01        &          $2.10 \pm 0.24$ \\ \hline
     T0560       &    2L02        &          $18.00 \pm 0.49$ \\ \hline
     T0569       &    2KWY        &          $20.01 \pm 0.69$  \\ \hline
\end{tabular}
\end{center}

\end{table}

>>>>

\author{Arijit~Roy}
\author{Alberto~Perez}
\author{Ken~A.~Dill}
\author{Justin~L.~MacCallum}
\affil{Laufer Center for Physical and Quantitative Biology\\
    Stony Brook University\\
    Stony Brook, NY 11794-5252.}

\title{Predicting the conformational preferences of proteins using a physics-based free energy
method}

@@

Protein conformation and associated conformational changes are important in biology. The mechanism
of enzymes, the function of structural proteins, and the binding of proteins to DNA or other
proteins all depend on protein conformations and the relative free energies of different
conformational states. Proteins can undergo conformational changes for a variety of reasons,
including ligand binding, change of pH or other conditions \cite{Meirovitch2007}, \cite{Chipot2007},
\cite{Jorgensen2004}. The process of protein folding also depends on the free energies of different
conformational states. The native structure is at a free energy minimum \cite{Anfinsen1973}, while
the kinetics are determined by the free energies of various on- and off-pathway intermediates.

However, the theoretical calculation of conformational free energy changes is difficult when the
states of interest are very different \cite{Meirovitch2007}. In these cases, the timescales involved
in such conformational transitions may be too long to simulate directly. For example, in protein
folding the free energy landscape is rough \cite{Dill1997}, \cite{Dill2008}, and the molecule can
become trapped in an energy well during conformational sampling. Even with advanced methods like
replica exchange molecular dynamics, the three dimensional structure may be trapped in a local
minima for a considerable time. Sometimes this problem can be circumvented by defining a pathway or
reaction coordinate connecting the states of interest. The free energy along this reaction
coordinate can then be determined using methods such as umbrella sampling \cite{Torrie1977},
targeted MD \cite{}, or thermodynamic integration \cite{Tironi1994}. However, accurately sampling
this reaction coordinate may still be difficult and require a large computationl effort. In the case
of protein folding, it is often difficult to even devise a meaningful reaction coordinate that could
connect the unfolded and folded states efficiently \cite{}.

The calculation of protein conformational free energy has been successfully attempted by a number of
groups \cite{Meirovitch2007,Ytreberg2006,Zheng2008}. In recent years, methods such as the reference
system method \cite{Ytreberg2006}, decativated morphing \cite{Park2008}, orthogonal space random
walk \cite{Zheng2008}, and the confinement method \cite{Tyka2006,Cecchini2009} have been used to
calculate free energy differences.

In this work we explore confinement method\cite{Tyka2006,Cecchini2009}, which relies on the fact
that the free energy difference is a state function and thus is independent of the sampling path.
This approach uses a thermodynamic cycle \ref{fig:method} where first protein conformational entropy
of the two end-states is frozen out by applying a series of restraints to the system. Next, the free
energy difference between the two highly-restrained harmocic systems is obtained using a normal mode
calculation.

Previously, this method was applied to calculate the conformational free energy of a 16 amino acid
residue peptide, known as BHP.  We have first reproduced that known result. Next, we have applied
this method to larger proteins in two different case studies. In the first, we attempt to pick the
most native like structures from a set of models produced during the Critical Assessment of
Structure Prediction (CASP) \cite{} experiment. In the second case study, we examine the
conformational preferences of a series of chameleon proteins designed by Orban and co-workers
\cite{} that can switch between two completely different folds with only small changes in amino acid
sequence.

@@

\subsection{Our implementation of the confiment method reproduces results from the literature}

As a first step, we performed several control experiments to verify that our implementation of the
confinement method produces results compatible with previous calculations reported in the
literature.

The method has previously been applied to a 16 amino acid residue $\beta$-hairpin from protein G,
known as BHP \cite{Cecchini2009}. We calculated the free energy difference between two different
conformations of the peptide: (1) the native conformation, called bhp1, with a two-stranded
$\beta$-sheet; and (2) a conformation, called bhp3, which has a three-stranded $\beta$-sheet.
Analysis of long (4 $\mu$s) equilibrium simulations \cite{Cecchini2009,Krivov2004}  shows that bhp1
is the more favorable configuration by 1.8 kcal/mol. Using the confinement method, we obtain a value
1.7 kcal/mol, which is in good agreement with the equilibrium simulations and with previous
calculations using the confinement method \cite{Cecchini2009}.

\subsection{The confiment method predicts that experimental structures have lower free energies than
computer-generated models}

Previous applications of the confiment method have focused on relatively short peptides, up to 17
residues in length. In this work we apply the confinement method to larger proteins. As a first
test, we calculated the free energies of models submitted by predictors during the CASP9 experiment
(often referred to as decoys) and compared them with the calculated free energy of the corresponding
experimentally determined structure from the Protein Data Bank (PDB). As
Table~\ref{table:casp_control} shows, in 5 out of 6 cases, the confinement method assigns a lower
free energy to the experimentally determined structure than to any of the decoys---as would be
expected. Although differentiating between experimentally determined structures and computer
generated predictions is not a stringent test of a scoring method---for example, examining the
residue-residue packing can easily distinguish between the two [ref Rosetta Holes]---the results
none the less serve as a useful "sanity check".

\begin{table}
\begin{center}
\caption{The confinement method assigns a more favorable free energy to the experimentally
    determined structure than to computer-generated predictions. For each target, we examined as
    many as five predictions submitted by CASP participants. We report the free energy difference
    between the most favorable decoy and the experimentally determined structure. Positive
    $\Delta\Delta G$ values indicate that the experimental structure is predicted to be more
favorable than any of the decoys.}
\label{table:casp_control}
\begin{tabular}{l l l}\hline
    CASP Target  & PDB Identifier & $\Delta\Delta G_{native \to best decoy}$ (kcal/mol) \\ \hline
     T0531       &    2KJX        &          $11.15 \pm 0.70$ \\ \hline
     T0538       &    2L09        &          $-3.00 \pm 0.47$ \\ \hline
     T0540       &    3MX7        &          $16.94 \pm 0.49$ \\ \hline
     T0559       &    2L01        &          $2.10 \pm 0.24$ \\ \hline
     T0560       &    2L02        &          $18.00 \pm 0.49$ \\ \hline
     T0569       &    2KWY        &          $20.01 \pm 0.69$  \\ \hline
\end{tabular}
\end{center}
\end{table}
